Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2015-11-09 docs/ms_sageAbundance.Rmd

Sagebrush (*Artemisia tridentata*) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].  SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015]. Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014]. In warmer parts of its range, this could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014]. In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014]. A large-scale spatially explicit population model for sagebrush driven by interannual climate would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.

@@

The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2).
Despite previous evidence that spring temperature strongly regulates population- and individual-level sagebrush performance [@Schlaepfer2012; @Schlaepfer2014], we found little evidence for a temperature effect (Fig. 2).
Our estimates for temperature effects broadly overlapped zero, but they did trend positive.
Thus, we did find some support for the idea that sagebrush will respond to positively to temperature increases in this colder part of the sagebrush range.
As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
Although our estimates of climate effects were small, they are in agreement with previous studies, improving our confidence that our model captured a realistic link between sagebrush growth and climate.

@@

Conclusion
----------
We have introduced a new approach to fitting and simulating population models at large spatial extents.
As methodological and conceptual advancements are made we will be able to apply our approach over even larger spatial extents, and at higher spatial resolution.
Working at large spatial extents, it will become feasible to use spatially-explicit weather data products to fit and simulate the models.
Given the dependence of forecasts on the weather projections used, we urge ecologists to adopt methods from weather forecasters to up- and down-weight GCMs based on study area-specific reliability.
Lastly, it may also be important to include the effects of species interactions [@Clark2014].
Fortunately, our spatio-temporal modeling framework is very flexible.
It can be used for any spatially-explicit time series of species-specific estimates of plant cover or density.
The core process model (Eq. 2) can take any form; meaning that additional covariates can easily be included or an entirely different functional form can be used depending on the ecological context.
As large scale time series data sets become more available, we anticipate the approach we developed here to fill the gap in spatial scales between species distribution models and local-scale demographic population models.

>>>>

Sagebrush (*Artemisia tridentata*) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].  SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015]. Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014]. In warmer parts of its range, this could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014]. In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014]. A large-scale spatially explicit population model for sagebrush driven by interannual climate would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.

@@

The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2) which agrees with previous research showing that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013]. The posterior distribution for the effect of spring temperature on sagebrush cover broadly overlapped zero but was slightly positive (Fig. 2).  Warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].  As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.

@@

While we presented projections for sagebrush cover at the end of this century, one important feature of our aproach is that it could also produce short-term forecasts for sagebrush cover as well.  For instance this apprach could be used to model the effects of a multi-year regional drought on this species (cite Debinski et al. 2010).  Validating spatial population models against short-term predictions would give modelers a way to assess the long-term forecasts produced by the model.  It would also provide opportunities for refinement of the model itself.  This cycle of prediction, validation and refinement is missing from most population-level forecasts of the effects of climate change currently produced [@Petchey2015 prediction paper -- is there a better reference for this idea?].

Conclusion
----------
We have introduced a new approach to fitting and simulating population models at large spatial extents with plant population data derived from state of the art remote sensing.
As more species-level remote sensing datasets become available and computing power increases this approach will be applicable to a wider number of species and even larger spatial extents. Future modeling could include the effects of non-climate drivers--including the effects of species interactions and disturbance [@Clark2014].  For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford and Lauenroth  2009 ].  Fortunately, our spatio-temporal modeling framework could easily be extended to model additional species and dynamic processes as the data become availabe. The approach we have developed here fills an important gap in spatial scales between species distribution models and local-scale demographic population models.
